The stopping power of hydrogen in hafnium and \\ the importance of relativistic $4f$ electrons

### 2020-06-02 sp_hafnium.tex
latest changes

%=======================================================================
%=======================================================================
\begin{abstract}
 The stopping power of protons through Hf foil has been studied both 
experimentally and theoretically. The measurements were performed at 
the Laboratory of Accelerators and X-Ray Diffraction in Lisbon by 
using the transmission method on self-supporting stopping material. 
 The overall uncertainty of around $5\%$ was established over the 
protons energy range  (0.6-2.5) MeV. The theoretical developments 
involved fully relativistic atomic structure calculations for Hf, which 
required the solution of the Dirac equation. The shell-wise local plasma 
approximation (SLPA) was used to describe the energy transferred to the 
bound $1s$-$4f$ electrons, and the outer four electrons were considered 
as a free electron gas (FEG). We found the relativistic description of 
the $4f$-shell and the screening between $4f$ and $5p$ electrons to be 
decisive around the stopping maximum. Present theoretical and 
experimental results are in excellent agreement in the energy region 
of the new measurements. However, our theoretical stopping 
cross sections show substantial differences with the most used 
semi-empirical models (SRIM2013 and ICRU-49) at intermediate to low 
energies. Our calculations suggest the stopping maximum to be higher and 
shifted to lower energies than these previous predictions. Future 
measurements below 100 keV would be necessary for a better understanding 
of the stopping power around the maximum and below.
\end{abstract}

@@

So far, only one experimental work has been published regarding the 
stopping power cross section of pure hafnium for protons~\cite{Sirotinin}, 
while more attention has been recently given to studies involving 
hafnium oxide due to its practical use~\cite{Abril,Behar,Primetzhofer,Roth}. 
It is well known that significant attention has been paid in recent 
years to transition metal-oxides such as HfO$_2$ because of their 
potential as alternative gate dielectrics to replace SiO$_2$ for the 
future generation of nano-electronics with less than 45 nm gate 
length~\cite{Choi,Robertson}. Some important physical properties of the 
above mentioned metal-oxide films depend on their thickness, which is 
often measured by using Rutherford Backscattering 
Spectrometry~\cite{Alfassi01,Tesmer01}. This method relies heavily on 
the determination of the stopping power of ion beams in the material of 
interest.

@@

%-----------------------------------------------------------------------
\begin{figure}[!t]
\centering
\includegraphics[width=9cm]{Fig01.eps}
\caption{RBS spectrum for $E_{\mathrm{avg}}=921.1$ keV protons on 
the hafnium sample, which is subsequently used to determine the energy 
loss in the foil.}
\label{F01}
\end{figure}

@@

%-----------------------------------------------------------------------
\subsection{Accelerator and scattering Chamber}
The procedure used in this work to obtain stopping power data is 
essentially the same as described in Ref.~\cite{Miranda01}. The present 
measurements were made at the IST/LATR (Laboratory of Accelerators and 
X-Ray Diffraction) in Lisbon. This facility uses a 2.5 MV Van de Graaff 
accelerator to deliver $^1$H$^+$ primary ion beams through a series of 
electrostatic lenses and collimators onto a thin Au/SiO$_2$ sample, 
which is used as a scattering center. This sample was placed in the 
center of an RBS/C scattering chamber, where a high vacuum (pressure of 
$\sim$10$^{-6}$ Torr) was maintained during the measurements. The beam 
current on the sample was kept at around 5.0 nA to attain sufficient 
statistics in each particle spectrum. By using a beam spot of about 
1.0 mm in diameter, a solid angle of 11.4 msr was attained. The overall 
energy resolution (FWHM) of the detection system was about 15 keV 
relative to 5.486 MeV alpha particles from a $^{241}$Am source.

%-----------------------------------------------------------------------
\subsection{Target}
The stopping material under analysis was a hafnium foil with a nominal 
thickness of 1.0 $\mu$m and 99.95\% purity, which was supplied by Lebow 
Company~\cite{Lebow}. However, a more precise thickness value was 
achieved by measuring the energy loss of alpha particles coming from a 
calibrated ($^{239}$Pu, $^{241}$Am, $^{244}$Cm) source. From the alpha 
spectra with and without the Hf foil interposed, the characteristic 
energy shift $\delta$E was measured and then combined with the stopping 
power for 5.486 MeV alphas on hafnium (55.69 eV/$10^{15}$ at/cm$^2$) 
found in Ref.~\cite{Ziegler01} to obtain an areal density of 
($4.13 \pm 0.21$)$\times 10^{19}$ at/cm$^2$, which corresponds to a 
thickness of $0.920\pm0.046 \mu$m.

%-----------------------------------------------------------------------
\subsection{Energy loss measurement}
Once the beam impinges on the Au/SiO$_2$ sample, protons are 
backscattered towards a Si surface barrier detector located at 
140$^{\circ}$ relative to the initial beam direction. Fig.~\ref{F01} 
shows two particle spectra, where the ion energies $E_2$ and $E_1$ are 
associated with a placed and removed hafnium sample, respectively. Both 
energy distributions were fitted by Gaussian functions to obtain the 
mean energy and width (FWHM) of the peaks \cite{Sun01}, and from the 
difference between these two peak positions in the spectrum, the total 
energy loss $\Delta E = (E_1 - E_2)$ in the foil was calculated. As 
established in previous studies \cite{Miranda01,Damache02}, the 
experimental stopping power cross sections $\varepsilon (E) $ are 
determined at some mean energy $E_{\mathrm{avg}}$ by measuring the ion 
energy losses $\Delta E$ within the investigated Hf foil, which has a 
mean thickness denoted by $\Delta x$. In this way, only when the energy 
loss fraction $\Delta E/E_{\mathrm{avg}}$ across the Hf foil is not 
exceeding 20\%, it is possible to define the stopping cross section by 
\cite{Raisanen01,Schulz01}:
\begin{equation}\label{eq:stcross}
 \varepsilon(E)=\frac{S(E)}{N}=-\frac{dE}{N\,dx}\approx-\frac{\Delta E}{N\Delta x},
\end{equation}
where $N$ denotes the atomic number density (atoms cm$^{-3}$) of the 
material under study. When this condition was not fulfilled, a small 
correction to the mean energies $E_{\mathrm{avg}}$ was applied in order 
to account for the non-linear dependence on ion energy of stopping 
powers \cite{Chilton,Rajatora}.

@@

%----------------------------------------------------------------------
\begin{figure*}[!t]
\centering
\includegraphics[width=11.cm]{bindener.eps}
\caption{(a) Binding energies of Hf. Present relativistic and available 
non-relativistic~\cite{badnell97} values are given with filled symbols. 
Experimental measurements for solids~\cite{williams1995} are depicted 
with open circles. (b) Corresponding relative errors with respect to 
experimental data.}
\label{Binding_E}
\end{figure*}
%------------------------------------------------------------------------

@@

More detail of the theoretical binding energies is given in 
Fig.~\ref{Binding_E}~(b), where relative errors with respect to the 
experimental values are shown. This figure shows clearly that the 
relativistic corrections are critical to describing the atomic structure 
of hafnium, even for the outer shells. It turns out that the errors 
committed in the non-relativistic calculations of the inner shell 
orbitals propagate, through the Hartree-Fock approximation, to the outer 
shells. The importance of fully relativistic calculations for the outer 
shells has already been noted for Au, Pb, Bi, and W~\cite{mon09}.

For the contribution of bound electrons to the total stopping cross 
sections, the SLPA considers independent contributions of each subshell. 
Our relativistic binding energies present spin-orbit split. However, 
in total stopping power, where the initial state of the excited electron 
is not measured, the quantum uncertainty in energy $\Delta E$ melts this 
split. The criterion $\Delta E\Delta t\geq\hbar/2$ merges the energies 
$E_{nl+}-E_{nl-}$ for sufficiently small values of $\Delta t$ (the 
collisional mean-time). In fact, at sufficiently high impact velocity, 
we can expect all target electrons to respond together to the ion 
passage~\cite{lindhard53,chu72}. Following previous works~\cite{mon09},
the collisional time is estimated as $\Delta t\approx\langle r_i\rangle/v$, 
with $\langle r_i\rangle$ and $v$ being the orbital mean radio and 
impact velocity, respectively. In the case of hafnium, we found that for 
every sub-shell of electrons, the spin-orbit split is unresolved in the 
energy region this sub-shell contributes. Therefore, the $nl$-electrons 
should be considered together, responding to the ion passage as a single 
gas of electrons with density $\delta_{nl}(r)$ and mean binding energy 
$E_{nl}$. This feature is vital within the SLPA calculations because 
it accounts for the screening among electrons of the \textit{same} 
binding energy. For example, the $4f_{-}$ and $4f_{+}$ of Hf can only 
be resolved for impact energies $E<0.05$ keV, but the contribution of 
$4f$ to the total stopping is negligible for $E<40$ keV. Moreover, the 
$5p$ and $4f$ electrons of Hf are very close in energy 
($\Delta E_{5p-4f} \approx 1$ a.u.~\cite{mendez2019}), and they react 
together at impact energies $E>40$ keV (inter-shell screening). As 
already mentioned, at higher energies, inter-shell screening is possible 
for other subshells (i.e., $4p$-$4d$ for impact energies above 0.9 MeV), 
but its weight in the total stopping is minor for deeper shells.

@@

In Fig.~\ref{slpa4f}, we display the present theoretical stopping cross
section of Hf for protons using the relativistic wave functions 
and binding energies, but with and without the $5p$-$4f$ screening. 
We show the FEG and bound electron contributions separately and the 
total stopping as the addition of both of them. The minimum energy for 
plasmon excitation was estimated at approximately $37$ keV. We used the 
non-perturbative SPCC model for impact energies $E \leq 37$~keV, and the 
perturbative ML calculation above this value. Bound $1s$-$4f$ electrons 
(relativistic wave functions and binding energies) are calculated with 
the SLPA and shown separately in Fig.~\ref{slpa4f} with and without the 
$4f$-$5p$ screening. Below $\sim 40$ keV, the difference between both 
calculations is negligible. Considering $5p$-$4f$ electrons 
as a single group of 20 electrons with screening among them gives lower 
stopping values than the addition of the separate $5p$ and $4f$ 
contributions. Notice that this shell correction can be considered 
naturally within a many-electron model, such as the SLPA.

@@

Fig.~\ref{F03} synthesizes the results of the present work. The 
agreement between the present theoretical results and the new 
measurements displayed in Table~\ref{table01} is excellent. Present 
measurements using the transmission method are in good agreement with 
the previous data by Sirotinin~\cite{Sirotinin}, which were measured 
in backscattering geometry. Our theoretical approach also agrees with 
the data by Sirotinin~\cite{Sirotinin}, except for the lowest energy 
measurement at 80 keV. We have also included in this figure the 
theoretical curves from the CasP5.2 code by Grande and 
Schiwietz~\cite{Grande,casp52} and from the DPASS code by Sigmund and 
Schinner~\cite{DPASS20}, both available online. Furthermore, we 
incorporated the semi-empirical results from SRIM-2013~\cite{Ziegler01} 
and the ICRU49 tables~\cite{ICRU49}. Interestingly, our full theoretical 
curve differs from SRIM-2013 for impact energies below 100 keV. We 
obtain a stopping maximum of approximately 
$40\times 10^{-15}$ eV cm$^2$/atom at 65 keV. Instead, following the 
up-to-now only set of data \cite{Sirotinin}, SRIM-2013 suggests a lower 
stopping maximum at an impact energy of 115 keV. The stopping maximum is 
a susceptible region for any full theoretical description without 
parameters, and this is quite visible in a linear-scale plot like 
Fig.~\ref{F03}. However, the impact energy for the maximum seems to 
agree between our curve and DPASS, although it is $10 \%$ below. 
Instead, CasP a maximum value of the stopping power $10 \%$ above our 
value and a completely different shape at lower energies. It is worth 
mentioning that our model gives similar results using the experimental 
value $r_S=2.07$ a.u. rather than the theoretical one $r_S=2.14$ a.u.,
with the stopping maximum at the same impact energy but $4\%$ higher.

Future experiments for impact energies below 100 keV would be essential 
to complete this study on the stopping of Hf for protons. In the energy 
region $80-500$ keV, only the measurements by Sirotinin's 
group~\cite{Sirotinin} from 1984 are available. SRIM-2013 includes these 
values in the code fitting, so it does not represent a test for this 
data. Undoubtedly, new measurements are needed in three specific 
energy regions: below 500 keV, around the stopping maximum 
(i.e., $50-200$ keV), and at low impact energies.

@@

To the best of our knowledge, these are the first theoretical 
calculations of stopping in Hf to take into account relativistic effects 
in the atomic structure and screening among electrons in a consistent 
way, from very low to high impact energies. Future experiments for 
impact energies around the stopping maximum and in the low energy region 
would be essential to have a better understanding of the stopping of 
protons in hafnium.

>>>>

%=======================================================================
%=======================================================================
\begin{abstract}
 The stopping power of protons through Hf foil has been studied both 
experimentally and theoretically. The measurements were performed at 
the Laboratory of Accelerators and X-Ray Diffraction in Lisbon by 
using the transmission method on self-supporting stopping material. 
 The overall uncertainty of around $5\%$ was established over the 
protons energy range  (0.6-2.5) MeV. The theoretical developments 
involved fully relativistic atomic structure calculations for Hf, which 
required the solution of the Dirac equation. The shell-wise local plasma 
approximation (SLPA) was used to describe the energy transferred to the 
bound $1s$-$4f$ electrons, and the outer four electrons were considered 
as a free electron gas (FEG). We found the relativistic description of 
the $4f$-shell and the screening between $4f$ and $5p$ electrons to be 
decisive around the stopping maximum. Present theoretical and 
experimental results are in very good agreement in the energy region 
of the new measurements. However, our theoretical stopping 
cross sections show substantial differences with the most used 
semi-empirical models (SRIM2013 and ICRU-49) at intermediate to low 
energies. Our calculations suggest the stopping maximum to be higher and 
shifted to lower energies than these previous predictions. Future 
measurements around the maximum and below would be necessary for a better understanding 
of the stopping power of hafnium.
\end{abstract}

@@

So far, only one experimental work has been published regarding the 
stopping power cross section of pure hafnium for protons~\cite{Sirotinin}, 
while more attention has been recently given to studies involving 
hafnium oxide due to its practical use~\cite{Abril,Behar,Primetzhofer,Roth}. 
It is well known that significant attention has been paid in recent 
years to transition metal-oxides such as HfO$_2$ because of their 
potential as alternative gate dielectrics to replace SiO$_2$ for the 
future generation of nano-electronics with less than 45 nm gate 
length~\cite{Choi,Robertson}. Some important physical properties of the 
above mentioned metal-oxide films depend on their thickness, which is 
often measured by using Rutherford Backscattering 
Spectrometry~\cite{Alfassi01,Tesmer01}. This method relies on 
the determination of both the scattering cross section and also the stopping power of ion beams in the material of 
interest.

@@

%-----------------------------------------------------------------------
\begin{figure}[!t]
\centering
\includegraphics[width=9cm]{Fig01.eps}
\caption{RBS spectrum for $E_{\mathrm{avg}}=921.1$ keV protons on 
hafnium sample which is subsequently used to determine the energy loss 
in the foil.}
\label{F01}
\end{figure}

@@

%-----------------------------------------------------------------------
\subsection{Accelerator and scattering Chamber}
The procedure used in this work to obtain stopping power data is 
essentially the same as described in Ref.~\cite{Miranda01}. The present 
measurements were made at the IST/LATR (Laboratory of Accelerators and 
X-Ray Diffraction) in Lisbon. This facility uses a 2.5 MV Van de Graaff 
accelerator to deliver $^1$H$^+$ primary ion beams with a precision better 
than $\pm 0.5$ keV through a series of electrostatic lenses and collimators 
onto a thin Au/SiO$_2$ sample, which is used as a scattering center. The Hf 
foil was mounted on a movable target holder and placed inside a RBS/C scattering 
chamber to allow energy measurements of the direct beam and the beam transmitted 
through the sample without breaking the high vacuum ($\sim$10$^{-6}$ Torr) inside 
the scattering chamber. The beam current on the sample was kept at around 5.0 nA 
to attain sufficient statistics in each particle spectrum. By using a beam spot 
of about 1.0 mm in diameter, a solid angle of 11.4 msr was attained. The overall 
energy resolution (FWHM) of the detection system was about 15 keV relative to 5.486 
MeV alpha particles from a $^{241}$Am source.

%-----------------------------------------------------------------------
\subsection{Target}
The stopping material under analysis was a hafnium foil with a nominal 
thickness of 1.0 $\mu$m and 99.95\% purity, which was supplied by Lebow 
Company~\cite{Lebow}. A more precise thickness value was achieved by measuring 
the energy loss of alpha particles coming from a calibrated 
($^{239}$Pu, $^{241}$Am, $^{244}$Cm) source. From the alpha 
spectra with and without the Hf foil interposed, the characteristic 
energy shift $\delta$E was measured and then combined with the stopping 
power for 5.486 MeV alphas on hafnium (55.69 eV/$10^{15}$ at/cm$^2$) 
found in Ref.~\cite{Ziegler01} to obtain an areal density of 
($4.13 \pm 0.21$)$\times 10^{19}$ at/cm$^2$, which corresponds to a 
thickness of $0.920\pm0.046 \mu$m. Target non-uniformity was investigated through 
systematic measurements (at five different points over the sample area) of the energy 
loss of alpha particles from the same radioactive source. The uncertainties originating
from the non-uniformity of the sample was $\sim 2.5$\%. However, the primary source of
uncertainty related to target thickness comes from estimates in the SRIM database for 
alphas on hafnium ($\sim$ 4\%). Additionally, we consider estimates coming from surface 
roughness ($\sim$ 1\%) and possible impurities ($\sim$ 1\%) in the foil; and finally, 
statistical uncertainty ($\sim$ 0.6\%) related to the gaussian fits used to determine 
the energy loss of alphas through the Hf target.
%-----------------------------------------------------------------------
\subsection{Energy loss measurement}
Once the beam impinges on the Au/SiO$_2$ sample, protons are 
backscattered towards a Si surface barrier detector located at 
140$^{\circ}$ relative to the initial beam direction. Fig.~\ref{F01} 
shows two particle spectra, where the ion energies $E_2$ and $E_1$ are 
associated with a placed and removed hafnium sample, respectively. Both 
energy distributions were fitted by Gaussian functions to obtain the 
mean energy and width (FWHM) of the peaks \cite{Sun01}, and from the 
difference between these two peak positions in the spectrum, the total 
energy loss $\Delta E = (E_1 - E_2)$ in the foil was calculated. As 
established in previous studies \cite{Miranda01,Damache02}, the 
experimental stopping power cross sections $\varepsilon (E) $ are 
determined at some mean energy $E_{\mathrm{avg}}$ by measuring the ion 
energy losses $\Delta E$ within the investigated Hf foil, which has a 
mean thickness denoted by $\Delta x$. In this way, only when the energy 
loss fraction $\Delta E/E_{\mathrm{avg}}$ across the Hf foil is not 
exceeding 20\%, it is possible to define the stopping cross section by 
\cite{Raisanen01,Schulz01}:
\begin{equation}\label{eq:stcross}
 \varepsilon(E)=\frac{S(E)}{N}=-\frac{dE}{N\,dx}\approx-\frac{\Delta E}{N\Delta x},
\end{equation}
where $N$ denotes the atomic number density (atoms cm$^{-3}$) of the 
material under study. When this condition was not fulfilled, a small 
correction to the mean energies $E_{\mathrm{avg}}$ was applied in order 
to account for the non-linear dependence on ion energy of stopping powers 
\cite{Chilton,Rajatora}. The uncertainty ($\sim$ 0.7\%) in the measured energy 
loss $\Delta E$ of protons in the hafnium sample is mainly related to the 
statistical uncertainty found in the gaussian fits mentioned above. If this value 
is combined with the $\sim$ 4.9\% uncertainty in target thickness, then a 
$\sim$ 5.0\% uncertainty in the measured cross section is obtained.

@@

%----------------------------------------------------------------------
\begin{figure*}[!t]
\centering
\includegraphics[width=11.cm]{bindener.eps}
\caption{(a) Binding energies of Hf. Present relativistic and available 
non-relativistic~\cite{badnell97} values are given with filled symbols. 
Experimental measurements for solids~\cite{williams1995} are depicted 
with open circles. (b) Corresponding relative errors respect to 
experimental data.}
\label{Binding_E}
\end{figure*}
%------------------------------------------------------------------------

@@

More detail about the theoretical binding energies is given in 
Fig.~\ref{Binding_E}~(b), where relative errors with respect to the 
experimental values are shown. This figure shows clearly that the 
relativistic corrections are critical to describe the atomic structure 
of hafnium, even for the outer shells. It turns out that the errors 
committed in the non-relativistic calculations of the inner shell 
orbitals propagate, through the Hartree-Fock approximation, to the outer 
shells. 
The non-relativistic $4f$ binding energy is four times the experimental one. Such an incorrect value leads to the underestimation of the $4f$-ionization and shifts  the threshold to higher energies.  
The importance of fully relativistic calculations for the outer 
shells has already been noted for Au, Pb, Bi, and W~\cite{mon09}.

For the contribution of bound electrons to the total stopping cross 
sections, the SLPA considers independent contributions of each subshell. 
Our relativistic binding energies present spin-orbit split. However, 
in total stopping power, where the initial state of the excited electron 
is not measured, the quantum uncertainty in energy $\Delta E$ melts this 
split. The criterion $\Delta E\Delta t\geq\hbar/2$ merges the energies 
$E_{nl+}-E_{nl-}$ for sufficiently small values of $\Delta t$ (the 
collisional mean-time). In fact, at sufficiently high impact velocity, 
we can expect all target electrons to respond together to the ion 
passage~\cite{lindhard53,chu72}. Following previous works~\cite{mon09},
the collisional time is estimated as $\Delta t\approx\langle r_i\rangle/v$, 
with $\langle r_i\rangle$ and $v$ being the orbital mean radio and 
impact velocity, respectively. In the case of hafnium, we found that for 
every sub-shell of electrons, the spin-orbit split is unresolved in the 
energy region this sub-shell contributes. Therefore, the $nl$-electrons 
should be considered together, responding to the ion passage as a single 
gas of electrons with density $\delta_{nl}(r)$ and mean binding energy 
$E_{nl}$. This feature is vital within the SLPA calculations because 
it accounts for the screening among electrons of the \textit{same} 
binding energy. For example, the $4f_{-}$ and $4f_{+}$ of Hf can only 
be resolved for impact energies $E<0.05$ keV, but the contribution of 
$4f$ to the total stopping is negligible for $E<40$ keV. Moreover, the 
$5p$ and $4f$ electrons of Hf are very close in energy 
($\Delta E_{5p-4f} \approx 1$ a.u.~\cite{mendez2019}), and they react 
together at impact energies $E>40$ keV (inter-shell screening). As 
already mentioned, at higher energies, inter-shell screening is also possible 
for deeper subshells %(i.e., $4p$-$4d$ for impact energies above 0.9 MeV), 
but their weight in the total stopping is minor.

@@

In Fig.~\ref{slpa4f}, we display the present theoretical stopping cross
section of Hf for protons using the relativistic wave functions 
and binding energies, but with and without the $5p$-$4f$ screening. 
We show the FEG and bound electron contributions separately and the 
total stopping as the addition of both of them. The minimum energy for 
plasmon excitation was estimated at approximately $37$ keV. We used the 
non-perturbative SPCC model for impact energies $E \leq 37$~keV, and the 
perturbative ML calculation above this value. Bound $1s$-$4f$ electrons 
(relativistic wave functions and binding energies) are calculated with 
the SLPA and shown separately in Fig.~\ref{slpa4f} with and without the 
$4f$-$5p$ screening. Below $\sim 40$ keV, the difference between both 
calculations is negligible. Considering $5p$-$4f$ electrons 
as a single group of 20 electrons with screening among them gives lower 
stopping values than the addition of the separate $5p$ and $4f$ 
contributions. Notice that this shell correction can only be considered self-consistently within a many-electron model, such as the SLPA.

@@

Fig.~\ref{F03} synthesizes the results of the present work. The 
agreement between the present theoretical results and the new 
measurements displayed in Table~\ref{table01} is excellent. Present 
measurements using the transmission method are in good agreement with 
the previous data by Sirotinin~\cite{Sirotinin}, which were measured 
in backscattering geometry. Our theoretical approach also agrees with 
the data by Sirotinin~\cite{Sirotinin}, except for the lowest energy 
measurement at 80 keV. We have also included in this figure the 
theoretical curves from the CasP5.2 code by Grande and 
Schiwietz~\cite{Grande,casp52} and from the DPASS code by Sigmund and 
Schinner~\cite{DPASS20}, both available online. Furthermore, we 
incorporated the semi-empirical results from SRIM-2013~\cite{Ziegler01} 
and the ICRU49 tables~\cite{ICRU49}. Interestingly, our full theoretical 
curve differs from SRIM-2013 for impact energies below 100 keV. We 
obtain a stopping maximum of approximately 
$40\times 10^{-15}$ eV cm$^2$/atom at 65 keV. Instead, following the 
up-to-now only set of data \cite{Sirotinin}, SRIM-2013 suggests a lower 
stopping maximum at an impact energy of 115 keV.

The stopping maximum is 
a very sensitive region for any full theoretical description, and this is quite visible in a linear-scale plot like 
Fig.~\ref{F03}. However, the impact energy for the maximum seems to 
agree between our curve and DPASS, although it is $10 \%$ below. 
Instead, CasP maximum is $10 \%$ above ours, but 
has a completely different shape at lower energies. It is worth 
mentioning that our model gives similar results using the experimental 
value $r_S=2.07$ a.u. rather than the theoretical one $r_S=2.14$ a.u.,
with the stopping maximum at the same impact energy but $4\%$ higher. Future experiments would be important for a more complete understanding of this case, mainly for proton energies around the stopping maximum (i.e. $30-300$ keV) and also below $25$ keV, in the region where a linear dependence with the velocity is expected.

@@

To the best of our knowledge, these are the first theoretical 
calculations of stopping in Hf that cover from very low to high impact energies, taking into account relativistic effects in the atomic structure and screening among electrons in a consistent 
way. Future experiments for 
impact energies around the stopping maximum and in the low energy region 
would be essential to have a better understanding of the stopping of 
protons in hafnium.
